Expert-In-The-Loop Causal Discovery: Iterative Model Refinement Using Expert Knowledge

### 2025-02-05 paper/main.tex
Updates

\begin{enumerate}
	\item \textbf{Continuous:} We train a model, $ E_X: x \sim
		\bm{z} $. The residuals are then computed by taking the difference
		between the true and the predicted values using $ E_X $. 
		$$ R_{x_i} = x_i - E_X(\bm{z}_i) $$
	\item \textbf{Ordinal:} We start by training a probability estimator, $
		p_X: x \sim \bm{z} $, and then use the estimated probabilities, 
		$ \hat{p}_X(x) $ to compute the residuals:
		$$ R_{x_i} = \hat{p}_X(X < x_i) - \hat{p}_X(X > x_i) $$
	\item \textbf{Categorical:} We again start by training a probability
		estimator $ p_X: X \sim \bm{Z} $, and obtain probability
		estimates $ \hat{p}_X: p_X(\bm{Z}) $. Next, we dummy encode the
		categorical variable and then compute the residuals as follows: 
		$$ R_X = x_i - \hat{p}(\bm{z}_i) $$
\end{enumerate}

@@

\begin{enumerate}
	\item \textbf{Bounded: } $ \rho $ is bounded between $ 0 $ (no
		association) to $ 1 $ (perfect linear association). This bounded
		nature simplifies the interpretation and allows for easy 
		comparisons across multiple values.
	\item \textbf{Independent of Sample Size: } The measure remains consistent
		regardless of the sample size ensuring reliability across 
		studies with different sample sizes.
	\item \textbf{Invariant to the Dummy Encoding: } We use dummy encoding 
		to compute the residual matrix for categorical variables. Since
		canonical correlations identify linear combinations across
		columns, this measure is invariant to the specific dummy
		encoding scheme used.
	\item \textbf{Equivalent to Partial Correlations for Continuous variables}
		Canonical correlations is equivalent to Pearson's correlation coefficient
		when both $ X $ and $ Y $ are continuous. Hence, $ \rho $ generalizes
		partial correlations to mixed data types.
	\item \textbf{Equivalent to polychoric and polyserial correlation for
			ordinal and continuous variables: }
		Under the assumption that the ordinal variable is generated by
		discretizing an underlying gaussian continuous variable, this
		effect size is equivalent to polychoric and polyserial
		correlation. As both of them recover the Pearson correlation
		coefficient.
	\item \textbf{Related to Cram\'er's V: } 
\end{enumerate}

@@

Using this mixed data partial association measure, we now propose a method for
construcing DAGs while incorporating expert knowledge iteratively. Given a
potentially empty DAG $ G $ and a dataset $ D $, a p-value threshold of $
\alpha $ and a measure of association threshold of $ \beta $, the construction
procedure works as follows:

\begin{enumerate}
	\item For each pair of variables $ X $ and $ Y $ in the model:
		\begin{itemize}
			\item If there is an edge between $ X $ and $ Y $, remove the edge to obtain the DAG $ \underline{G} $, and 
				compute the measure of association $ \phi(X, Y \rvert \underline{G}_{pa}(X) \cup \underline{G}_{pa}(Y)) $
			\item If there is no edge between $ X $ and $ Y $, compute the measure of association $ \phi(X, Y \rvert G_{pa}(X) \cup \underline{G}_{pa}(Y) $.
		\end{itemize}
	\item If for any pair $ X $ and $ Y $, if there is an edge between them with p-value $ > \alpha $ and $ \phi < \beta $, remove the edge between them.
	\item For the remaining $ X $ and $ Y $ pairs with p-value $ \geq \alpha $, sort them by $ \beta $. Select the pair with highest $ \beta $, ask the user to specify the direction between $ X $ and $ Y $ and add the edge to $ G $.
\end{enumerate}

@@

\begin{figure}[t!]
	\centering
	\includegraphics[scale=0.4]{../code/plots/web_tool_full.png}
	\caption{A screenshot of the web tool for constructing the model. Users
		can upload their dataset after which the tool creates an empty
		graph and shows all pair of variables which are associated in
		the model using undirected red edges with the strength of
		association represented using edge width. Users can then add
		edges to the model (shown in green). Unnecessary edges are shown in 
		black.}
	\label{fig:web}
\end{figure}

@@

\section{Conclusions}
Points:
\begin{itemize}
	\item As researchers prefer to anyways build models by hand, this is like a companion tool to help them in deciding the method.
	\item Solves all the $ 3 $ issues outlined in the introduction.
	\item Compared to other expert knowledge specification methods, users don't need to guess what information then need to provide as exact questions are asked. Also enables us to ask an LLM for it.
\end{itemize}

>>>>

\begin{enumerate}
	\item \textbf{Continuous:} We train a model, $ E_X: x \sim
		\bm{z} $. The residuals are then computed by taking the difference
		between the true and the predicted values using $ E_X $. 
		$$ R_{x_i} = x_i - E_X(\bm{z}_i) $$
	\item \textbf{Ordinal:} We start by training a probability estimator, $
		p_X: x \sim \bm{z} $, and then use the estimated probabilities, 
		$ \hat{p}_X(x) $ to compute the residuals:
		$$ R_{x_i} = \hat{p}_X(X < x_i) - \hat{p}_X(X > x_i) $$
	\item \textbf{Categorical:} We again start by training a probability
		estimator $ p_X: X \sim \bm{Z} $, and obtain probability
		estimates $ \hat{p}_X: p_X(\bm{Z}) $. Next, we dummy encode the
		categorical variable and then compute the residuals as follows: 
		$$ R_{x_i} = x_i - \hat{p}(\bm{z}_i) $$
\end{enumerate}

@@

\begin{enumerate}
	\item \textbf{Bounded: } $ \rho $ is bounded between $ 0 $ (no
		association) to $ 1 $ (perfect linear association). This bounded
		nature simplifies the interpretation and allows for easy 
		comparisons across multiple values.
	\item \textbf{Independent of Sample Size: } The measure remains consistent
		regardless of the sample size ensuring reliability across 
		studies with different sample sizes.
	\item \textbf{Invariant to the Dummy Encoding: } We use dummy encoding 
		to compute the residual matrix for categorical variables. Since
		canonical correlations identify linear combinations across
		columns, this measure is invariant to the specific dummy
		encoding scheme used.
	\item \textbf{Equivalent to Partial Correlation}
		When both $ X $ and $ Y $ are continuous variables, our measure
		is equivalent to the absolute value of partial correlation.
	\item \textbf{Equivalent to polychoric and polyserial correlation for
			ordinal and continuous variables: }
		Under the assumption that the ordinal variable is generated by
		discretizing an underlying gaussian continuous variable, this
		effect size is equivalent to polychoric and polyserial
		correlation. As both of them recover the Pearson correlation
		coefficient.
	\item \textbf{Related to Cram\'er's V: } 
\end{enumerate}

@@

Using this mixed data measure of association, we now propose a method for
iteratively constructing DAGs while incorporating expert knowledge. This
approach is based on implied CI based testing where we use the measure of
association to rank the failing CI tests. This ranking allows us to prioritize
adding edges between pair of variables that have the most severe problems.

Given a DAG $ G $ (potentially empty), a dataset $ D $, a p-value threshold $
\alpha $ and a measure of association threshold $ \beta $, our construction
procedure works as follows:

\begin{enumerate}
	\item For each pair of variables $ X $ and $ Y $ in $ G $:
		\begin{itemize}
			\item If there is an edge between $ X $ and $ Y $, remove the edge to obtain the DAG $ G' $, and compute the measure of association $ \phi(X, Y \rvert G'_{pa}(X) \cup G'_{pa}(Y)) $
			\item If there is no edge between $ X $ and $ Y $, compute the measure of association $ \phi(X, Y \rvert G_{pa}(X) \cup \underline{G}_{pa}(Y) $.
		\end{itemize}
	\item If for any pair $ X $ and $ Y $, if there is an edge between them with p-value $ > \alpha $ and $ \phi < \beta $, remove the edge between them.
	\item For the remaining $ X $ and $ Y $ pairs with p-value $ \geq \alpha $, sort them by $ \beta $. Select the pair with highest $ \beta $, ask the user to specify the direction between $ X $ and $ Y $ and add the edge to $ G $.
\end{enumerate}

@@

\begin{figure}[t!]
	\centering
	\includegraphics[scale=0.4]{../code/plots/web_tool_full.png}
	\caption{A screenshot of the web tool for constructing the model. Users
		can upload their dataset after which the tool creates an empty
		graph and shows all pair of variables which are associated in
		the model using undirected red edges with the strength of
		association represented using edge width. Users can then
		iteratively add edges to the model (shown in green) while
		deciding the edge orientation based on domain knowledge.
		Unnecessary edges are shown in black.}
	\label{fig:web}
\end{figure}

@@

\section{Conclusions}

Points:
\begin{itemize}
	\item As researchers prefer to anyways build models by hand, this is like a companion tool to help them in deciding the method.
	\item Solves all the $ 3 $ issues outlined in the introduction.
	\item Compared to other expert knowledge specification methods, users don't need to guess what information then need to provide as exact questions are asked. Also enables us to ask an LLM for it.
\end{itemize}
